Phylogenetic comparative methods in the lme4-verse

### 2019-04-10 example.rmd
done

---
title: "Phylogenetic comparative methods in the lme4-verse"
author: Michael Li and Ben Bolker
date: "`r format(Sys.time(), '%H:%M %d %B %Y')`"
output:
    html_document:
    mathjax: default
---

```{r pkgs,message=FALSE, warning=FALSE}
library(ape)
library(lme4)
library(Matrix)
library(MCMCglmm)
library(MASS) 
library(pez)
library(glmmTMB)
library(dplyr)
library(coda)
library(lattice)
library(broom)
library(dotwhisker)
```

- The standard problem of *phylogenetic comparative methods* is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree - for example, regression analyses of body temperature as a function of body size for animals within a clade
- More generally, we can frame this in the usual GLMM way as
$$
\begin{split}
y & \sim D(\mu,\phi) \\
\mu & = g^{-1}(\eta) = g^{-1}(X \beta + Z b) \\
b & \sim \textrm{MVN}(0,\Sigma)
\end{split}
$$
where the part that makes it specifically phylogenetic is that $\Sigma$ captures the *phylogenetic correlation*. The PC is the correlation among observations due to relatedness; recently diverged taxa have higher correlation than more anciently diverged taxa. In the extreme case of a *star phylogeny* (all taxa diverged from each other simultaneously at some point in the past) the phylogenetic correlation collapses to a diagonal matrix and we get back to the simple, uncorrelated regression.

Various P(G)LMM (phylogenetic [generalized] linear mixed model] approaches have been proposed. Many depend on Pagel's lambda transformation, which gives the correlation matrix a particularly simple form (but has been [criticized](http://www.carlboettiger.info/2013/10/11/is-it-time-to-retire-pagels-lambda.html) ...)

An alternative approach is to model the phylogenetic correlation as a *Gaussian process*. In particular, suppose that the evolutionary process is a Brownian motion (an almost certainly incorrect/oversimplified model of evolution, but one that many phylogenetic methods are built on). In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past. If we set up the $Z$ matrix appropriately, we can model everything with a sequence of *independent* errors, rather than having to do fancy things to impose a correlation structure on the random effects.

## Nuts and bolts: from a phylogeny to a $Z$ matrix for the GP

```{r setup}
library(ape)
set.seed(101)
r <- makeNodeLabel(rtree(4))
plot(r,show.node.label=TRUE)
```

Information in a `phylo` object is contained in the *edge matrix*:

>    edge: a two-column matrix of mode numeric where each row represents
          an edge of the tree; the nodes and the tips are symbolized
          with numbers; the tips are numbered 1, 2, ..., and the nodes
          are numbered after the tips. For each row, the first column
          gives the ancestor.

```{r edge}
t(r$edge)
```

and a list of *edge lengths*

```{r edgelen}
r$edge.length
```

Inspecting this tree, we can figure out (see `$tip.label` and `$node.label` for label-to-number correspondences):

- tips are 1-4, nodes are 5-7
- tip 1 (`t1`) involves branches 2 (6 $\to$ 1) and 1 (5 $\to$ 6).
- tip 2 (`t3`) involves branches 3 (6 $\to$ 2) and 1 (5 $\to$ 6)
- tip 3 (`t2`) involves branches 5 (7 $\to$ 3) and 4 (5 $\to$ 7)
- tip 4 (`t4`) involves branches 6 (7 $\to$ 4) and 4 (5 $\to$ 7)

So, for example, we can say that the 'error' value corresponding to tip 1 is
$\ell_1 b_1 + \ell_2 b_2$,  where $\ell_i$ is the (square root of??) the branch length
and the $b_i$ are independent, homoscedastic Normal variates. Alternately, the $Z$ matrix is

$$
\begin{pmatrix}
\ell_1 & \ell_2 &   0     & 0 & 0 & 0 \\
\ell_1 &    0   &  \ell_3 & 0 & 0 & 0 \\
0 & 0 &  0  & \ell_4 & \ell_5 & 0 \\
0 & 0 &  0  & \ell_4 & 0 & \ell_6
\end{pmatrix}
$$
where $\ell_i$ is the length of the $i^\textrm{th}$ branch,
so that the species effects are $Z b$.

If we can build the corresponding $Z$ matrix, then we can insert it in
the `lme4` modular model-fitting process (see `?modular`).

Here's a (probably not very efficent) way to construct the Z matrix.
(There must be a way to not walk the tree multiple times from every tip ...

```{r phylo.to.Z}
phylo.to.Z <- function(r,stand=FALSE){
  ntip <- length(r$tip.label)
  Zid <- Matrix(0.0,ncol=length(r$edge.length),nrow=ntip)
  nodes <- (ntip+1):max(r$edge)
  root <- nodes[!(nodes %in% r$edge[,2])]
  for (i in 1:ntip){
    cn <- i  ## current node
    while (cn != root){
      ce <- which(r$edge[,2]==cn)   ## find current edge
      Zid[i,ce] <- 1   ## set Zid to 1
      cn <- r$edge[ce,1]            ## find previous node
    }
  }
  V <- vcv(r)
  # V <- V/max(V)
  sig <- exp(as.numeric(determinant(V)["modulus"])/ntip)
  # sig <- det(V)^(1/ntip)
  Z <- t(sqrt(r$edge.length) * t(Zid))
  if(stand){Z <- t(sqrt(r$edge.length/sig) * t(Zid))}
  rownames(Z) <- r$tip.label
  colnames(Z) <- 1:length(r$edge.length)
  return(Z)                                  
}
phylo.to.Z(r)
```

(This could benefit from the repeated-entry sparse matrix class that Steve Walker wrote.)

On the other hand, it only takes a few seconds to run for a 200-species phylogeny (see below).

## constructing a GP PGLMM with lme4: machinery

"All" we need to do is (1) call `(g)lFormula`, with a formula that includes a `(1|phylo)` term, to build the basic (wrong) structure; (2) modify the `reTrms` component of the structure appropriately; (3) go through the rest of the modular procedure for building a (G)LMM.

```{r phylocode}
source("lme4_phylo_setup.R")
```

## glmmTMB fits: nuts and bolts

`glmmTMB` can be deconstructed in a similar way. In fact, we can re-use a lot of the
machinery. Being able to use `glmmTMB` means we can use a broader range of distributions,
zero-inflation, etc. (machinery below assumes phylogenetic structure only in the
conditional distribution). This is also a little clunky, some adjustment on the `glmmTMB`
side might make it a bit easier.

```{r tmbfit0}
source("glmmTMB_phylo_setup.R")
```

## Example

### get data

>>>>

---
title: "Example"
author: "Michael Li and Ben Bolker"
output:
  pdf_document: default
  mathjax: default
  html_document: null
---

We will reproduce the examples in [chapter 11](http://www.mpcm-evolution.com/practice/online-practical-material-chapter-11/chapter-11-1-simple-model-mcmcglmm) of Garamszegi (ed.) using lme4 and glmmTMB.

To fit we need a random effect in the formula that includes a `(...|phylo)` term, to build the basic random effect structure multipled by the phyloZ matrix (see vignette for more details on phyloZ matrix).

```{r pkgs,message=FALSE, warning=FALSE}
library(ape)
library(Matrix)
library(lme4)
library(MCMCglmm)
library(MASS) 
library(pez)
library(glmmTMB)
library(dplyr)
library(coda)
library(lattice)
library(broom)
library(dotwhisker)
```

```{r phylocode}
source("lme4_phylo_setup.R")
source("glmmTMB_phylo_setup.R")
```

### Get data

### 2019-04-10 vignette.rmd
done

---
title: "Phylogenetic comparative methods in the lme4-verse"
author: Michael Li and Ben Bolker
date: "`r format(Sys.time(), '%H:%M %d %B %Y')`"
output:
    html_document:
    mathjax: default
---

```{r pkgs,message=FALSE, warning=FALSE}
library(ape)
library(lme4)
library(Matrix)
library(MCMCglmm)
library(MASS) 
library(pez)
library(glmmTMB)
library(dplyr)
library(coda)
library(lattice)
library(broom)
library(dotwhisker)
```

- The standard problem of *phylogenetic comparative methods* is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree - for example, regression analyses of body temperature as a function of body size for animals within a clade
- More generally, we can frame this in the usual GLMM way as
$$
\begin{split}
y & \sim D(\mu,\phi) \\
\mu & = g^{-1}(\eta) = g^{-1}(X \beta + Z b) \\
b & \sim \textrm{MVN}(0,\Sigma)
\end{split}
$$
where the part that makes it specifically phylogenetic is that $\Sigma$ captures the *phylogenetic correlation*. The PC is the correlation among observations due to relatedness; recently diverged taxa have higher correlation than more anciently diverged taxa. In the extreme case of a *star phylogeny* (all taxa diverged from each other simultaneously at some point in the past) the phylogenetic correlation collapses to a diagonal matrix and we get back to the simple, uncorrelated regression.

Various P(G)LMM (phylogenetic [generalized] linear mixed model] approaches have been proposed. Many depend on Pagel's lambda transformation, which gives the correlation matrix a particularly simple form (but has been [criticized](http://www.carlboettiger.info/2013/10/11/is-it-time-to-retire-pagels-lambda.html) ...)

An alternative approach is to model the phylogenetic correlation as a *Gaussian process*. In particular, suppose that the evolutionary process is a Brownian motion (an almost certainly incorrect/oversimplified model of evolution, but one that many phylogenetic methods are built on). In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past. If we set up the $Z$ matrix appropriately, we can model everything with a sequence of *independent* errors, rather than having to do fancy things to impose a correlation structure on the random effects.

@@

So, for example, we can say that the 'error' value corresponding to tip 1 is
$\ell_1 b_1 + \ell_2 b_2$,  where $\ell_i$ is the (square root of??) the branch length
and the $b_i$ are independent, homoscedastic Normal variates. Alternately, the $Z$ matrix is

@@

Here's a (probably not very efficent) way to construct the Z matrix.
(There must be a way to not walk the tree multiple times from every tip ...

```{r phylo.to.Z}
phylo.to.Z <- function(r,stand=FALSE){
  ntip <- length(r$tip.label)
  Zid <- Matrix(0.0,ncol=length(r$edge.length),nrow=ntip)
  nodes <- (ntip+1):max(r$edge)
  root <- nodes[!(nodes %in% r$edge[,2])]
  for (i in 1:ntip){
    cn <- i  ## current node
    while (cn != root){
      ce <- which(r$edge[,2]==cn)   ## find current edge
      Zid[i,ce] <- 1   ## set Zid to 1
      cn <- r$edge[ce,1]            ## find previous node
    }
  }
  V <- vcv(r)
  # V <- V/max(V)
  sig <- exp(as.numeric(determinant(V)["modulus"])/ntip)
  # sig <- det(V)^(1/ntip)
  Z <- t(sqrt(r$edge.length) * t(Zid))
  if(stand){Z <- t(sqrt(r$edge.length/sig) * t(Zid))}
  rownames(Z) <- r$tip.label
  colnames(Z) <- 1:length(r$edge.length)
  return(Z)                                  
}
phylo.to.Z(r)
```

(This could benefit from the repeated-entry sparse matrix class that Steve Walker wrote.)

@@

## constructing a GP PGLMM with lme4: machinery

"All" we need to do is (1) call `(g)lFormula`, with a formula that includes a `(1|phylo)` term, to build the basic (wrong) structure; (2) modify the `reTrms` component of the structure appropriately; (3) go through the rest of the modular procedure for building a (G)LMM.

```{r phylocode}
source("lme4_phylo_setup.R")
```

## glmmTMB fits: nuts and bolts

`glmmTMB` can be deconstructed in a similar way. In fact, we can re-use a lot of the
machinery. Being able to use `glmmTMB` means we can use a broader range of distributions,
zero-inflation, etc. (machinery below assumes phylogenetic structure only in the
conditional distribution). This is also a little clunky, some adjustment on the `glmmTMB`
side might make it a bit easier.

```{r tmbfit0}
source("glmmTMB_phylo_setup.R")
```

## Example

### get data

From [chapter 11](http://www.mpcm-evolution.com/practice/online-practical-material-chapter-11/chapter-11-1-simple-model-mcmcglmm) of Garamszegi (ed.): data are [here](http://mpcm-evolution.com/OPM/Chapter11_OPM/data.zip)

```{r getdat}
if (!file.exists("data/phylo.nex")) {
    dir.create("data")
    download.file("http://mpcm-evolution.com/OPM/Chapter11_OPM/data.zip",
                  dest="data/OPM_ch11_data.zip")
    setwd("data")
    untar("OPM_ch11_data.zip")
    setwd("..")
}
phylo <- read.nexus("data/phylo.nex")
```

Compute appropriate $Z$ matrix up front, to measure speed (also
reusable in a few places below):

```{r calc_phyloZ,cache=TRUE}
system.time(phyloZ <- phylo.to.Z(phylo))
```

## Result comparison with Gaussian example in chapter 11

```{r G_results}
datG <- read.table("data/data_simple.txt",header=TRUE)
datG$obs <- factor(seq(nrow(datG)))
datG <- datG %>% mutate(sp = phylo)
phylo_lmm_fit <- phylo_lmm(phen~cofactor+(1|sp)
  , data=datG
  , phylonm = "sp"
  , phylo = phylo
  , phyloZ=phyloZ
  , REML = TRUE
  , control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore")
)

print(summary(phylo_lmm_fit))

```

## Result comparison with Gaussian with repeated measures example in chapter 11

```{r GR_results}
datR <- read.table("data/data_repeat.txt",header=TRUE)
datR$obs <- factor(seq(nrow(datR)))
datR <- (datR 
  %>% mutate(sp = species
        , animals = phylo
        )
)
datR$spec_mean_cf <- sapply(split(datR$cofactor,datR$phylo),mean)[datR$phylo]
datR$within_spec_cf <- datR$cofactor-datR$spec_mean_cf
phylo_lmm_fit <- phylo_lmm(phen~spec_mean_cf+within_spec_cf+(1|sp) + (1|animals)
  , data=datR
  , phylonm = "sp"
  , phylo = phylo
  , phyloZ=phyloZ
  , REML = FALSE
  , control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore")
)

print(summary(phylo_lmm_fit))

```

## Result comparison with non-Gaussian example in chapter 11

```{r P_results}
dat <- read.table("data/data_pois.txt",header=TRUE)
dat$obs <- factor(seq(nrow(dat)))

dat <- dat %>% mutate(sp=phylo) 
phylo_glmm_fit <- phylo_glmm(phen_pois~cofactor+(1|sp)+(1|obs)
  , data=dat
  , phylonm = "sp"
  , family = poisson
  , phylo = phylo
  , phyloZ=phyloZ
  , control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore")
)

summary(phylo_glmm_fit)

```

>>>>

---
title: "Phylogenetic comparative methods in the lme4-verse"
author: "Michael Li and Ben Bolker"
output:
  pdf_document: default
  mathjax: default
  html_document: null
---

The goal of *phylogenetic comparative methods* is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree - for example, regression analyses of body temperature as a function of body size for animals within a clade
More generally, we can frame this in the usual GLMM way as
$$
\begin{split}
y & \sim D(\mu,\phi) \\
\mu & = g^{-1}(\eta) = g^{-1}(X \beta + Z b) \\
b & \sim \textrm{MVN}(0,\Sigma)
\end{split}
$$
where the part that makes it specifically phylogenetic is that $\Sigma$ captures the *phylogenetic correlation* (PC). The PC is the correlation among observations due to relatedness; recently diverged taxa have higher correlation than more anciently diverged taxa. In the extreme case of a *star phylogeny* (all taxa diverged from each other simultaneously at some point in the past) the phylogenetic correlation collapses to a diagonal matrix and we get back to the simple, uncorrelated regression.

Various P(G)LMM (phylogenetic [generalized] linear mixed model) approaches have been proposed. Many depend on Pagel's lambda transformation, which gives the correlation matrix a particularly simple form (but has been [criticized](http://www.carlboettiger.info/2013/10/11/is-it-time-to-retire-pagels-lambda.html) ...)

An alternative approach is to model the phylogenetic correlation as a *Gaussian process* (GP). In particular, suppose that the evolutionary process is a Brownian motion process (an almost certainly incorrect/oversimplified model of evolution, but one that many phylogenetic methods are built on). In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past. If we set up the $Z$ matrix appropriately, we can model everything with a sequence of *independent* errors, rather than having to impose a correlation structure on the random effects.

@@

So, for example, we can say that the 'error' value corresponding to tip 1 is
$\ell_1 b_1 + \ell_2 b_2$,  where $\ell_i$ is the square root of the branch length
and the $b_i$ are independent, homoscedastic Normal variates. Alternately, the $Z$ matrix is

@@

Here's a (probably not very efficient) way to construct the Z matrix.
(There must be a way to not walk the tree multiple times from every tip ...)

```{r phylo.to.Z}
phylo.to.Z <- function(r,stand=FALSE){
  ntip <- length(r$tip.label)
  Zid <- Matrix(0.0,ncol=length(r$edge.length),nrow=ntip)
  nodes <- (ntip+1):max(r$edge)
  root <- nodes[!(nodes %in% r$edge[,2])]
  for (i in 1:ntip){
    cn <- i  ## current node
    while (cn != root){
      ce <- which(r$edge[,2]==cn)   ## find current edge
      Zid[i,ce] <- 1   ## set Zid to 1
      cn <- r$edge[ce,1]            ## find previous node
    }
  }
  V <- vcv(r)
  sig <- exp(as.numeric(determinant(V)["modulus"])/ntip)
  Z <- t(sqrt(r$edge.length) * t(Zid))
  if(stand){Z <- t(sqrt(r$edge.length/sig) * t(Zid))}
  rownames(Z) <- r$tip.label
  colnames(Z) <- 1:length(r$edge.length)
  return(Z)                                  
}
phylo.to.Z(r)
```

@@

To fit in lme4 and glmmTMB, we need random effect terms in the formula that includes a `(...|phylo)`. Using the lme4 machinery, it will build the appropriate random effect structure multiplied by the phyloZ matrix.
`glmmTMB` can be deconstructed in a similar way. In fact, we can re-use a lot of the
machinery. Being able to use `glmmTMB` means we can use a broader range of distributions,
zero-inflation, etc. (machinery below assumes phylogenetic structure only in the
conditional distribution). This is also a little clunky, some adjustment on the `glmmTMB`
side might make it a bit easier.
